Understanding the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations

### 2022-04-17 manuscript-Ghana/manuscript-Ghana.Rmd
Adding the R markdown file

---
title: Short Paper
author:
  - name: Issoufou Liman Harou
    email: issoufoul@gmail.com
    affiliation: ICRAF-GOV, KU-ENV
    correspondingauthor: true
  - name: Julliet Inyele 
    email: inyelejuli@gmail.com
    affiliation: KU-GEO
  - name: Lalisa Duguma
    email: l.duguma@cgiar.org
    affiliation: ICRAF
address:
  - code: ICRAF-GOV
    address: 'Governance and Landscapes, World Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O.Box 30677-00100'
  - code: KU-ENV
    address: 'Department of Environmental Sciences, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100'
  - code: KU-GEO
    address: 'Department of Geography, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100'
abstract: |
  Remote sensing analysis and narratives from Ghana reported evidence of localized mangrove degradations. Despite the importance of mangrove forests for Ghana, the coverage of earth observation satellites, and computing facilities, accurate country-wide information on mangrove dynamics are lacking. The persistence of cloud cover is an important limiting factor for optical remote sensing in the coastal regions of Ghana. In this study, we identified 17 land cover categories and consider all available Landsat images acquired between 2000 and 2020 to study the dynamics of mangroves in Ghana. Considering the surface reflectance and brightness temperature bands, we harmonized the Landsat bands across sensors, used time series models to improve cloud detection and recovered the signal encoding the spectral behavior of land cover over time. Considering the land cover phenology, the shortages of Landsat images due to excessive cloud cover in the study area and the revisit time of Landsat satellite, we mosaiced the Landsat images within 30-days temporal windows. Based on the fitted image time series, we computed unsupervised classifications which we used along with ground control polygons to generate accurate land cover reference data. These were used as training and validation data to develop random forest classifiers for classification and change detection. The analysis of the root mean square error indicates that the accuracy of the fitted image time series increases with band wavelength and land cover seasonality. The accuracy assessment indicates an overall accuracy ranging between 0.94 and 0.95, and a Kappa ranging between 0.93 and 0.95 for the different land cover classifications. Mangroves have declined in some areas, whereas they have increased in others. Mangrove area, which have decrease over the last decade (2000- 2010), have increased during the previous decade (2010 - 2020). The gains experienced between 2010 and 2020, however, did not compensate for the lost the land cover experienced between 2000 and 2010. We identified four top most important land cover involved in mangroves conversion including forests with a high rate, croplands and natural vegetation mosaics and tree plantations with moderate rate and wetlands with a low rate. Better mangrove restoration and conservation policies are required to fully reverse the trend of mangrove decline in Ghana.
keywords: 
  - Landsat Time-series
  - Multitemporal
  - Cloud Mask
  - Land Cover Classification
  - Change Detection
  - Mangrove
  - Google Earth Engine
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: bibliography.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/remote-sensing-of-environment
csl: remote-sensing-of-environment.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
link-citations: yes
linkcolor: blue
header-includes:
  - \input{includes.tex}
note: Replication files are available on the author's Github account (http://github.com/Issoufou-Liman).
---

@@

if (!require("remotes")) {
  install.packages("remotes")
}
if (!require("rgee")) {
  remotes::install_github("r-spatial/rgee")
}
if(!require("tinytex")) {
  install.packages("tinytex")
  tinytex::install_tinytex()
}
## loading required packages ####
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  googledrive,
  googleCloudStorageR,
  rgee,
  tidyverse,
  grid,
  gridExtra,
  rgdal,
  raster,
  sf,
  ggspatial,
  kableExtra
)

@@

## Should local files with the same name be overwritten?
overwrite_file = TRUE

@@

source('source_files/ee/01_functions.R')
source('source_files/ee/02_commonSpecs.R')

if(!drive_folder_exists("Mangrove_Ghana_manuscript")){
  drive_mkdir("Mangrove_Ghana_manuscript")
}

ee_manage_create(paste0(asset_home, '/Mangrove_Ghana_manuscript'), asset_type = "Folder", quiet = FALSE)

@@

dir.create(file.path('source_files'))
dir.create(file.path('source_files/ee'))
dir.create(file.path('source_files/r'))
dir.create(file.path('ee_output'))
dir.create(file.path('ee_output/Ghana_waters_polygons'))
dir.create(file.path('data_files'))
dir.create(file.path('figures'))

```

<!-- ```{r ee-compute-01-unsupervised-classifications,include=FALSE,echo=FALSE,cache=FALSE} -->
<!-- # Compute and save an unsupervised classification and save it to ee asset -->

<!-- ##################/ Collection 1999 - 2020  ################### -->

<!-- # source('rgee/functions/functions_Ghana.R') -->
<!-- filteredLandsat =  collectLansatImages ('LSR', roiGhana, 1999, 2020, ee$List$sequence(1, 12), nDays, TRUE); -->

<!-- filteredLandsat = ee$ImageCollection(filteredLandsat); -->

<!-- # Some fixed variable for easy maintenance -->
<!-- rawBands = ee$List(list('blue', 'green', 'red', 'nir', 'swir1', 'swir2', 'temp', 'NDVI', 'NDWI')); -->

<!-- filledHarmonic = harmonic_fill(filteredLandsat, rawBands, harmonics); -->

<!-- # Plot the fitted model and the original data at the ROI. -->
<!-- filledBands = rawBands$map(ee_utils_pyfunc(function(band){ -->
<!--   return (ee$String(band)$cat('_')$cat('filled')); -->
<!-- })); -->

<!-- fittedBands = rawBands$map(ee_utils_pyfunc(function(band){ -->
<!--   return (ee$String(band)$cat('_')$cat('fitted')); -->
<!-- })); -->

<!-- theseBands = rawBands$cat(filledBands)$cat(fittedBands); -->

<!-- fitBands = ee$List(list('blue', 'green', 'red', 'nir', 'swir1', 'swir2'))$ -->
<!--   map(ee_utils_pyfunc(function(band){ -->
<!--     return (ee$String(band)$cat('_')$cat('filled')); -->
<!--   })); -->

<!-- fittedHarmonic = harmonic_fit(filledHarmonic, fitBands, harmonics, lagDays); -->

<!-- filledBands = fitBands$map(ee_utils_pyfunc(function(band){ -->
<!--   return (ee$String(band)$cat('_')$cat('filled')); -->
<!-- })); -->

<!-- fittedBands = fitBands$map(ee_utils_pyfunc(function(band){ -->
<!--   return (ee$String(band)$cat('_')$cat('fitted')); -->
<!-- })); -->

<!-- theseBands = fitBands$cat(filledBands)$cat(fittedBands); -->

<!-- ##################/ Unsupervised classification 2000 - 2002  ################### -->

<!-- classificationBands = ee$List(list('blue', 'green', 'red', 'nir', 'swir1', 'swir2'))$ -->
<!--   map(ee_utils_pyfunc(function(band){ -->
<!--     return (ee$String(band)$cat('_')$cat('filled')$cat('_')$cat('fitted')); -->
<!--   })); -->
<!-- fittedHarmonic_2000 = fittedHarmonic$select(classificationBands)$ -->
<!--   filterDate('2000-01-01', '2002-12-31'); -->

<!-- bandSeries_2000 = fittedHarmonic_2000$toBands(); -->

<!-- unsupervised_2000 = unsuperClassifyThis(bandSeries_2000, 17, AgScale, roiGhana, 8, 1000)#$rename('unsupervised_2000'); -->

<!-- assetId = paste0(asset_path, '/unsupervised_2000') -->
<!-- if(!asset_exists(assetId)){ -->
<!--   task_img <- ee_image_to_asset( -->
<!--     image = unsupervised_2000, -->
<!--     description = "unsupervised_2000", -->
<!--     assetId = assetId, -->
<!--     overwrite = FALSE, -->
<!--     region = roiGhana, -->
<!--     scale = AgScale, -->
<!--     maxPixels = 1e13 -->
<!--   ) -->
<!--   task_img$start() -->
<!--   ee_monitoring(task_img) -->
<!-- } -->
<!-- ##################/ Unsupervised classification 2010 - 2012  ################### -->

<!-- fittedHarmonic_2010 = fittedHarmonic$select(classificationBands)$ -->
<!--   filterDate('2010-01-01', '2012-12-31'); -->

<!-- bandSeries_2010 = fittedHarmonic_2010$toBands(); -->

<!-- unsupervised_2010 = unsuperClassifyThis(bandSeries_2010, 17, AgScale, roiGhana, 8, 1000)$rename('unsupervised_2010'); -->

<!-- assetId = paste0(asset_path, '/unsupervised_2010') -->
<!-- if(!asset_exists(assetId)){ -->
<!--   task_img <- ee_image_to_asset( -->
<!--     image = unsupervised_2010, -->
<!--     description = "unsupervised_2010", -->
<!--     assetId = assetId, -->
<!--     overwrite = FALSE, -->
<!--     region = roiGhana, -->
<!--     scale = AgScale, -->
<!--     maxPixels = 1e13 -->
<!--   ) -->
<!--   task_img$start() -->
<!--   ee_monitoring(task_img) -->
<!-- } -->
<!-- ##################/ Unsupervised classification 2018 - 2020  ################### -->

<!-- fittedHarmonic_2018 = fittedHarmonic$select(classificationBands)$ -->
<!--   filterDate('2018-01-01', '2020-12-31'); -->

<!-- bandSeries_2018 = fittedHarmonic_2018$toBands(); -->

<!-- unsupervised_2018 = unsuperClassifyThis(bandSeries_2018, 17, AgScale, roiGhana, 8, 1000)$rename('unsupervised_2018'); -->

<!-- assetId = paste0(asset_path, '/unsupervised_2018') -->
<!-- if(!asset_exists(assetId)){ -->
<!--   task_img <- ee_image_to_asset( -->
<!--     image = unsupervised_2018, -->
<!--     description = "unsupervised_2018", -->
<!--     assetId = assetId, -->
<!--     overwrite = FALSE, -->
<!--     region = roiGhana, -->
<!--     scale = AgScale, -->
<!--     maxPixels = 1e13 -->
<!--   ) -->
<!--   task_img$start() -->
<!--   ee_monitoring(task_img) -->
<!-- } -->
<!-- ``` -->

```{r ee-compute-02-referenceData,include=FALSE,echo=FALSE,cache=FALSE}
assetId <- paste0(asset_path, "/Ghana_field_data")
if(!asset_exists("Ghana_field_data", asset_path)){
  source("source_files/ee/03_referenceData.R")
  #### Unsupervised class identification and matching ####

  # The following images are made available only. This is for 2 raisins:
  # To ensure reproducibility
  # They are time consuming in computation
  unsupervised_2000 = ee$Image('users/Liman/210311_unsupervised2000_Ghana');
  unsupervised_2010 = ee$Image('users/Liman/210311_unsupervised2010_Ghana');
  unsupervised_2018 = ee$Image('users/Liman/210311_unsupervised2018_Ghana');
  # Map$addLayer(unsupervised_2000, list(palette = myPalette, min = 0, max = 16), 'unsupervised_2000', legend = FALSE);
  # Map$addLayer(unsupervised_2010, list(palette = myPalette, min = 0, max = 16), 'unsupervised_2010', legend = FALSE);
  # Map$addLayer(unsupervised_2018, list(palette = myPalette, min = 0, max = 16), 'unsupervised_2018', legend = FALSE);

  numPoints = 700
  # Class identification and matching
  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies = matches$
    stratifiedSample(
      numPoints = 250, # numPoints*3/8, 
      region = waterZone, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));
  # ee_print(waterBodies, 'waterBodies')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'waterBodies matches');
  # Map$addLayer(waterBodies, list(color = 'blue'), 'waterBodies pts');

  past = 11; p10 = 11; przt = 11;
  #  past = 9; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies_2 = matches$
    stratifiedSample(
      numPoints = 100, # numPoints*1/8, 
      region = waterZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));
  # ee_print(waterBodies_2, 'waterBodies_2')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'waterBodies_2 matches');
  # Map$addLayer(waterBodies_2, list(color = 'blue'), 'waterBodies_2 pts');

  ##############################
  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Non-fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));
  # ee_print(seaWater_1, 'seaWater_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(seaWater_1, list(color = 'green'), 'seaWater_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 1)
      );
    }));
  # ee_print(closeForests_1, 'closeForests_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'close forest');
  # Map$addLayer(closeForests_1, list(color = 'green'), 'closeForests_1 pts');

  past = 0; p10 = 0; przt = 0;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'open forests matches');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));
  # ee_print(openForests_1, 'openForests_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'openForests matches');
  # Map$addLayer(openForests_1, list(color = 'green'), 'openForests_1 pts');
  openForests_2 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = openForestZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));
  # ee_print(openForests_2, 'openForests_2')
  # Map$addLayer(openForests_2, list(color = 'green'), 'openForests_2 pts');
  openForests_3 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = openForestZone_3, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));
  # ee_print(openForests_3, 'openForests_3')
  # Map$addLayer(openForests_3, list(color = 'green'), 'openForests_3 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Woody savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 3)
      );
    }));
  # ee_print(woodySavannas_1, 'woodySavannas_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(woodySavannas_1, list(color = 'green'), 'woodySavannas_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 4)
      );
    }));
  # ee_print(savannas_1, 'savannas_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(savannas_1, list(color = 'green'), 'savannas_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 5)
      );
    }));
  # ee_print(closedShrubs_1, 'closedShrubs_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(closedShrubs_1, list(color = 'green'), 'closedShrubs_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open Shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 6)
      );
    }));
  # ee_print(openShrubs_1, 'openShrubs_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(openShrubs_1, list(color = 'green'), 'openShrubs_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Grasslands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 7)
      );
    }));
  # ee_print(grasslands_1, 'grasslands_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(grasslands_1, list(color = 'green'), 'grasslands_1 pts');

  #  past = 5; p10 = 5; przt = 5;
  past = 2; p10 = 6; przt = 15;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  wetlands_1 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = wetlandsZone_1,
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));
  # ee_print(wetlands_1, 'wetlands_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(wetlands_1, list(color = 'green'), 'wetlands_1 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));
  # ee_print(wetlands_2, 'wetlands_2')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(wetlands_2, list(color = 'green'), 'wetlands_3 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));
  # ee_print(wetlands_3, 'wetlands_3')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(wetlands_3, list(color = 'green'), 'wetlands_3 pts');

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Croplands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 9)
      );
    }));
  # ee_print(mosaic_1, 'mosaic_1')
  # Map$addLayer(matches, list(palette = list('white', 'red')), 'matches');
  # Map$addLayer(mosaic_1, list(color = 'green'), 'mosaic_1 pts');

@@

  task_vector$start()
  ee_monitoring(task_vector) 
}
```

>>>>

---
title: Understanding the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations
author:
  - name: Issoufou Liman Harou
    email: issoufoul@gmail.com
    affiliation: ICRAFGOV,KUENV
    correspondingauthor: true
  - name: Julliet Inyele 
    email: inyelejuli@gmail.com
    affiliation: KUGEO
  - name: Peter Minang  
    email: A.Minang@cgiar.org
    affiliation: ICRAFGOV
  - name: Lalisa Duguma
    email: l.duguma@cgiar.org
    affiliation: ICRAFGOV
address:
  - code: ICRAFGOV
    address: "Governance and Landscapes, World Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O.Box 30677-00100"
  - code: KUENV
    address: "Department of Environmental Sciences, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100"
  - code: KUGEO
    address: "Department of Geography, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100"
abstract: |
  Remote sensing analysis and narratives reported evidence of mangrove degradation at various spatial scale across Ghana. Despite the importance of mangrove forests, the coverage of earth observation satellites, and the availability of cloud computing platforms for remote sensing data analysis, accurate country-wide information on the dynamics of these mangrove forests is lacking. Cloud cover is an important limitation for optical remote sensing in Southern Ghana. To understand the dynamics of mangrove forests, we identified 17 land cover classes, harmonized the Landsat scenes across sensors, used time series analysis to improve cloud detection and recover the relevant signal encoding the temporal signature of land cover. Based on the seasonal behavior of the land cover classes, the persistence of cloud cover limiting the number of usable images and the temporal resolution of Landsat satellite, we mosaiced the images within 30 days of their acquisition dates. Using the fitted image time series, we computed unsupervised classifications which we used along with ground control polygons to generate training and validation data for land cover classifications and change detection. The analysis of the root mean square error indicates that the accuracy of the fitted image time series increases with band wavelength and land cover seasonality. The accuracy assessment indicates an overall accuracy between 0.94 and 0.95 and a Kappa between 0.93 and 0.95 for the different land cover classifications. Mangrove forests have declined in some areas, whereas they have increased in others. Mangrove area, which has decreased between 2000 and 2010, has increased between 2010 and 2020. The gains experienced between 2010 and 2020, however, did not compensate for the lost the land cover experienced between 2000 and 2010. We identified four top most important land cover involved in mangroves conversion including forests with a high rate, croplands and natural vegetation mosaics and tree plantations with moderate rate and wetlands with a low rate. Better mangrove restoration and conservation are required to fully reverse the trend of mangrove decline.
keywords: 
  - Google Earth Engine
  - Land Cover Classification
  - Cloud Cover
  - Mangrove
  - Change Detection
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: bibliography.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/remote-sensing-of-environment
csl: remote-sensing-of-environment.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
link-citations: yes
linkcolor: blue
header-includes:
  - \input{includes.tex}
note: Replication files are available on the author's Github account (http://github.com/Issoufou-Liman).
---

```{r,include=FALSE}
options(tinytex.engine='xelatex')
pdf.options(useDingbats = TRUE)
```

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=TRUE,out.extra='',fig.align='center',fig.pos='!htbp',cache=TRUE,echo=FALSE,warning=FALSE,message=FALSE)
```

@@

if (!require("remotes")) {
  install.packages("remotes")
}
if (!require("rgee")) {
  remotes::install_github("r-spatial/rgee")
}
if(!require("tinytex")) {
  install.packages("tinytex")
  tinytex::install_tinytex()
}
## loading required packages ####
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  googledrive,
  googleCloudStorageR,
  rgee,
  tidyverse,
  grid,
  gridExtra,
  tiff,
  rgdal,
  raster,
  sf,
  ggspatial,
  kableExtra
)
if (!require("gdalUtils")) install.packages("gdalUtils")

@@

## Should local files with the same name be overwritten?
overwrite_file = FALSE

@@

source('source_files/ee/01_functions.R')

if(!drive_folder_exists("Mangrove_Ghana_manuscript")){
  drive_mkdir("Mangrove_Ghana_manuscript")
}
ee_manage_create(paste0(asset_home, '/Mangrove_Ghana_manuscript'), asset_type = "Folder", quiet = FALSE)

@@

dir.create(file.path('source_files'))
dir.create(file.path('source_files/ee'))
dir.create(file.path('source_files/r'))
dir.create(file.path('ee_output'))
dir.create(file.path('ee_output/Ghana_waters_polygons'))
dir.create(file.path('data_files'))
dir.create(file.path('figures'))
dir.create(file.path('figures/tifs'))
dir.create(file.path('figures/pngs'))

source('source_files/ee/02_commonSpecs.R')
map(list('.shp', '.shx', 'prj', 'dbf', 'cpg'), function(i){
  ee_to_drive_to_local (
    ee_object = ee$FeatureCollection(list(ee$Feature(roiGhana))),
    drive_description = '~/Mangrove_Ghana_manuscript/roiGhana', 
    drive_folder = 'Mangrove_Ghana_manuscript', 
    region = roiGhana, 
    scale = AgScale,
    maxPixels = 1e13, 
    local_folder  = 'ee_output/roiGhana',
    overwrite = overwrite_file,
    ee_type = 'table',
    file_extension = i
  )
})
```

```{r ee-compute-02-referenceData,include=FALSE,echo=FALSE,cache=FALSE}
assetId <- paste0(asset_path, "/Ghana_field_data")
if(!asset_exists("Ghana_field_data", asset_path)){
  source("source_files/ee/03_referenceData.R")

  ## Unsupervised class identification and matching

  # The following images are made available only. This is for 2 raisins:
  # To ensure reproducibility
  # They are time consuming in computation
  unsupervised_2000 = ee$Image('users/Liman/210311_unsupervised2000_Ghana');
  unsupervised_2010 = ee$Image('users/Liman/210311_unsupervised2010_Ghana');
  unsupervised_2018 = ee$Image('users/Liman/210311_unsupervised2018_Ghana');
  numPoints = 700
  # Class identification and matching
  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies = matches$
    stratifiedSample(
      numPoints = 250, 
      region = waterZone, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

  waterBodies_2 = matches$
    stratifiedSample(
      numPoints = 100, 
      region = waterZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),
      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

  past = 11; p10 = 11; przt = 11;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Non-fresh water')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 0)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 1)
      );
    }));

  past = 0; p10 = 0; przt = 0;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches')

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

  openForests_2 = matches$
    stratifiedSample(
      numPoints = numPoints*2/5, 
      region = openForestZone_2, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

  openForests_3 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = openForestZone_3, 
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open forests')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 2)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Woody savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 3)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Savannas')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 4)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Close shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 5)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Open Shrublands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 6)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Grasslands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 7)
      );
    }));

    past = 2; p10 = 6; przt = 15;
  matches = unsupervised_2000$eq(past)$And(unsupervised_2010$eq(p10))$And(unsupervised_2018$eq(przt))$rename('matches');
  wetlands_1 = matches$
    stratifiedSample(
      numPoints = numPoints*1/5, 
      region = wetlandsZone_1,
      scale = AgScale, 
      classValues = list(0), 
      classPoints = list(0),

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Wetlands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 8)
      );
    }));

@@

      tileScale = 4, 
      geometries = TRUE
    )$map(ee_utils_pyfunc(function(ft){
      return (
        ft$set('class', 'Croplands')$
          set('matches', NULL)$
          set('past', past)$
          set('p10', p10)$
          set('present', przt)$
          set('landcover', 9)
      );
    }));

@@

  task_vector$start()
  ee_task_monitoring(task_vector) 
}
```
